Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-16 ThesisJP/Introduction/introduction.tex
Now, only VR and Feild simulations sections remains. And some iterations with correcting language.

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

@@

The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging, an effective processing rate of around 100 MFLOPS is required. This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.  Nevertheless, with modern GPUs, these levels of processing are finally available within a single card. When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

In \textbf{Paper\,II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper\,II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper\,II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly. Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally. The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm. This led us to the work of \textbf{Paper\,III}. 

@@

In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling. These errors are naturally reduced when the lateral sampling is increased. This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging. The reason could be ignorance or its possible negative impact on imaging frame rate. A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments. Fortunately, as shown in \textbf{Paper\,III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays). It also turns out that more matrix inversions are not needed either. Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on scan-grid data. %The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase. The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming. 

@@

With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms. For example to increase the precision of speckle and edge tracking. In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided. Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature. Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.

@@

One limitation of the algorithm described in \textbf{Paper\,IV} is its dependence on a successful tracking. If the tracking fails the proposed method will fail.  Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle. Even if the method successfully removes noise, this means that the rendering will become non-intuitive to interact with and unrealistic to watch. However, the method could still be used to clean up still images of standard cardiac views. The paper does also propose an interesting method for automatic tuning of rendering parameters. Figure 3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics. Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Figure 3d is performed. A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work. An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme is likely to be better suited for cardiac ultrasound imaging than a local opacity transfer function. A local opacity transfer function is just to extreme to be clinically robust. It would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity. It should also be possible to use this statistic as input to image filtering. An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces. In that way we can apply less filtering and higher opacity on edges. If this will be different than just looking at neighboring pixels in a smoothing filter is to be seen.

@@

The fifth paper stands out from the other papers by not dealing with an adaptive processing technique. It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on GPUs. Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence, for instance to achieve a certain pressure in a given area. The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables. Estimation of the mechanical and thermal index could also be improved. Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own. It has to be combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained. These are points we have tried to answer in both \textbf{Paper\,I}, \textbf{II},  \textbf{IV}, and \textbf{V}.  In \textbf{Paper\,V} the increased speed made an interactive ''Paint-program'' feasible. As of January 2014 the simulator has been download 200 times and counting. It also has close to a "whopping" 1000 views on youtube. 

>>>>

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision this is more than the worlds most powerful supercomputer could provide in 1998, and that was a room full of computers.

@@

The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging, an effective processing rate of hundreds of GFLOPS is required. This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second. Nevertheless, with modern GPUs, these levels of processing are finally available within a single card. When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

In \textbf{Paper\,II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper\,II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper\,II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly. %Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally. 
The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm. This led us to the work of \textbf{Paper\,III}. 

@@

In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling. These errors are naturally reduced when the lateral sampling is increased. This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging. The reason could be ignorance or its possible negative impact on imaging frame rate. A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments. Fortunately, as shown in \textbf{Paper\,III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays). It also turns out that more matrix inversions are not needed either. Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on polar-grid data. %The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase. The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming. 

@@

With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms. For example to increase the precision of speckle and edge tracking. In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided and the shift-invariance property is maintained. Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature. Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.

@@

One limitation of the algorithm described in \textbf{Paper\,IV} is its dependence on a successful tracking. If the tracking fails the proposed method will fail.  Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle. Even if the method successfully removes noise, this means that the rendering will become non-intuitive to interact with and unrealistic to watch. However, the method could still be used to clean up still images of standard cardiac views. The paper does also propose an interesting method for automatic tuning of rendering parameters. Fig.\,3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics. Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,3d in \textbf{Paper\,IV} is performed. A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work. An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme is likely to be better suited for cardiac ultrasound imaging than a local opacity transfer function. A local opacity transfer function is just to extreme to be clinically robust. In addition, it would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity. It should also be possible to use this statistic as input to image filtering. An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces. In that way we can apply less smoothing and higher opacity at tissue interfaces. If this will be different than just looking at neighboring pixels in an edge-preserving image filter is to be seen.

@@

The fifth paper stands out from the other papers by not dealing with an adaptive processing technique. It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on the GPU. Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence, for instance to achieve a certain pressure in a given area. The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables. Estimation of the mechanical and thermal index could then also be improved. Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own. It has to be combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained. These are points we have tried to answer in both \textbf{Paper\,I}, \textbf{II},  \textbf{IV}, and \textbf{V}.  In \textbf{Paper\,V} the increased speed made an interactive ''Paint-program'' feasible. As of January 2014 the simulator has been download 200 times and counting. It also has close to a "whopping" 1000 views on youtube. 
